---
title: "Crime Rates"
---

Follow along with the demo to learn how to calculate crime rates using R and the Tidyverse. We'll also introduce a new library, janitor, which helps make column names easier to work with. You should get started by loading the tidyverse in a new R block, then install and load janitor.

### Task 1: Load the tidyverse and install and load janitor
```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)
```


### Task 2: Load the crime dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv
```{r}
annual_crime_rate<-read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv")
```
### Task 3: Use janitor to clean up the column names

```{r}

annual_crime_rate <- annual_crime_rate %>%clean_names()
```

### Task 4: Calculate the total crime rate for each year (picking our unit!)

```{r}
totalRate<-annual_crime_rate %>% mutate(totalRate = (grand_total/population)*10000)
```

### Task 5: Calculate the violent and property crime rates for each year (same unit)
```{r}
annual_crime_rate <- annual_crime_rate %>%
  mutate(
    violent_crime_rate = (violent_crime_total / population) * 10000,
    property_crime_rate = (property_crime_total / population) * 10000)

```
### Task 6: Describe our results

Write a few sentences about what you found in the data. What years had the highest crime rates? What years had the lowest crime rates? Does the pattern match your expectations?

From 1999 going forward there has been a steady decrease in the total rate of crimes. The year with the lowest crime rate was 2020 with around 204 crimes/10,000 people. The year with the highest crime was 1980 with around 663 crimes/10,000 people. The following years with the highest crime rates were 1981, 1979, 1995, 1992 and 1991. I'm surprised because the population has been increasing and crime rates have decreased a lot since the highest reported years. I did hear that the 90's were the deadliest decade (according to the TV channel DNA, I think it's called) so it's not that surprising that there were a lot of crimes during that time period. I think before the 2000's it might've been easier to get away with crimes too without all of the technology that's used today, which led to a lot more crimes occuring. 


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

