---
title: "Final_Project"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
install.packages("janitor")
library(janitor)
```

```{r}
crime <- read_csv("https://data.lacity.org/api/views/2nrs-mtv8/rows.csv?accessType=DOWNLOAD")
```

```{r}
crime <- crime %>%clean_names()
```
Trials Errors and Lessons: 
I ask ChatGPT a few things for these ones so here is the link to all of that: https://chatgpt.com/share/682aa394-8abc-800f-8bda-4e3c2813753c 

Narrowing the data to a single number of crime in a certain category using totals: 

```{r}
crime %>%
  group_by(2020) %>%
  summarise(total_cases = n())
```
```{r}
crime %>%
  group_by(2020, vict_sex) %>%
  summarise(total_cases = n())
```
For the table above, the internet was pretty divided on what X (intersex, gender not listed, gender not included, non-binary...) meant but ChatGPT said X means unknown or unreported gender. The internet or ChatGPT did not know what H meant. 


Tried this to try and filter gender and type of crime but it didn't work. That's okay though, I'll just use the amount of crimes commited in 2020 total and look at that. I tried changing the year and crime and stuff but I don't know. 
```{r}
crime %>%
  filter(date_occ == 2021, crm_cd_desc == "BURGLARY") %>%
  group_by(vict_sex) %>%
  summarise(total_cases = n())

```
ChatGPT said to add the lubridate... because I'm asking for a specific year and the dates are listed as full dates which is why I'm not seeing anything. It still didn't work. I don't know, oh well. 
```{r}
library(dplyr)
library(lubridate)

crime %>%
  filter(date_occ == 2021, crm_cd_desc == "BURGLARY") %>%
  group_by(vict_sex) %>%
  summarise(total_cases = n())
```

Calculating Means and SD Values: 
```{r}
crime %>% summarize(mean=mean(vict_age), sd=sd(vict_age))
```
Asked ChatGPT how to narrow it down by year: 
```{r}
library(dplyr)
library(lubridate)

crime %>%
  filter(year(date_occ) == 2021) %>%
  summarize(mean_age = mean(vict_age, na.rm = TRUE),
            sd_age = sd(vict_age, na.rm = TRUE))
```
```{r}
str(crime$date_occ)
head(crime$date_occ)
```
Chat said these are ambigous formats of dates that makes it unrecognizable for lubridate. That makes sense why the earlier code wasn't working. 
```{r}
crime <- crime %>%
  mutate(date_occ = as.Date(date_occ, format = "%m/%d/%Y"))
```
This converts the dates. Now it should work...
```{r}
library(lubridate)

crime %>%
  filter(year(date_occ) == 2020) %>%
  summarize(mean_age = mean(vict_age, na.rm = TRUE),
            sd_age = sd(vict_age, na.rm = TRUE))
```
YAYY! Lets see if we can filter even further with gender. Chat said non-numeric values go in quotation marks.

```{r}
crime %>%
  filter(vict_sex == "F")%>%
 filter(vict_sex == "M")%>%
filter(year(date_occ) == 2020) %>%
  summarize(mean_age = mean(vict_age, na.rm = TRUE),
            sd_age = sd(vict_age, na.rm = TRUE))
```
Ooh. No okay. Well. I'll take the "M" catergory out and see. So I guess the below is the mean and sd for females in 2020. 
```{r}
crime %>%
  filter(vict_sex == "F")%>%
filter(year(date_occ) == 2020) %>%
  summarize(mean_age = mean(vict_age, na.rm = TRUE),
            sd_age = sd(vict_age, na.rm = TRUE))
```
Here is males then: 
```{r}
crime %>%
  filter(vict_sex == "M")%>%
filter(year(date_occ) == 2020) %>%
  summarize(mean_age = mean(vict_age, na.rm = TRUE),
            sd_age = sd(vict_age, na.rm = TRUE))
```
Okay. 

New Min/Max victim age with filter and within the new catergory: 
crime |> 
  filter(vict_age > 0) |> 
  summarize(min_age = min(vict_age), max_age = max(vict_age)) 
```{r}
min_age <- min(crime$vict_age, na.rm = TRUE)
max_age <- max(crime$vict_age, na.rm = TRUE)
cat("Victim ages range from", min_age, "to", max_age)

crime |> 
  filter(vict_age > 0) |> 
  summarize(min_age = min(vict_age), max_age = max(vict_age))
```
  That worked. Ages range from 2 to 120 years old. I'm going to try and filter it. 
  
```{r}
min_age <- min(crime$vict_age, na.rm = TRUE)
max_age <- max(crime$vict_age, na.rm = TRUE)
cat("Victim ages range from", min_age, "to", max_age)

crime |> 
  filter(vict_age > 0) |> 
  filter(year(date_occ)== 2020) %>%
  summarize(min_age = min(vict_age), max_age = max(vict_age))

```
```{r}

min_age <- min(crime$vict_age, na.rm = TRUE)
max_age <- max(crime$vict_age, na.rm = TRUE)
cat("Victim ages range from", min_age, "to", max_age)

crime |> 
  filter(vict_age > 0) |> 
  filter(year(date_occ)== 2020) %>%
  filter(vict_sex == "M")%>%
  summarize(min_age = min(vict_age), max_age = max(vict_age))
```
```{r}

min_age <- min(crime$vict_age, na.rm = TRUE)
max_age <- max(crime$vict_age, na.rm = TRUE)
cat("Victim ages range from", min_age, "to", max_age)

crime |> 
  filter(vict_age > 0) |> 
  filter(year(date_occ)== 2020) %>%
  filter(vict_sex == "F")%>%
  summarize(min_age = min(vict_age), max_age = max(vict_age))
```
  
```{r}

crime %>%
  filter(year(date_occ) == 2020) %>%
  filter(vict_age)%>%
  summarize(mean_age = mean(vict_age, na.rm = TRUE),
            sd_age = sd(vict_age, na.rm = TRUE)
```
Chat said to change it to this because I'm filtering numbers now.
```{r}
crime %>%
  filter(year(date_occ) == 2020) %>%
  summarize(mean_age = mean(vict_age, na.rm = TRUE),
            sd_age = sd(vict_age, na.rm = TRUE))
```

  Okay Cool. Updated Chat link just in case it didn't update the last one as I went: https://chatgpt.com/share/682aa394-8abc-800f-8bda-4e3c2813753c 
  
Tests I can run: 
I can find the total amount of women and men who were victims. 
Find the average age and gender of victims. 
I tried to do the glimpse crime, date range and age range but I have to look more into age range and what the glimpse of data actually does as I work on this today. 
I can find which days have the highest amount of crimes happening, who was the most targeted victim and then research what could have happened that day in that area. 
Which area has the most and least crime.

```{r}
glimpse(crime)
```
```{r}
colSums(is.na(crime))
```
```{r}
min_date <- min(crime$date_occ)
max_date <- max(crime$date_occ)
cat("Data ranges from", min_date, "to", max_date)

```
```{r}
min_age <- min(crime$vict_age, na.rm = TRUE)
max_age <- max(crime$vict_age, na.rm = TRUE)
cat("Victim ages range from", min_age, "to", max_age)

```
This age range lowkey does not sound right... 

Data Explorations: 
The source of this data is data.gov and it's crime data from 2020 to 2023. I think I was doing mental health or hate crime data before but I could not figure out how to import that data and it was taking too much time to figure out. Here is the link to that data: https://catalog.data.gov/dataset/crime-data-from-2020-to-present/resource/5eb6507e-fa82-4595-a604-023f8a326099l. 

The crime data provided has data from 2020 to now in L.A. The crimes range from stolen vehicles, vandalism, rape, theft, arson and more along with the area they were committed in. I can analyze who has the most crimes committed against them. The most important columns for this analysis are victim age and victim sex. I think I would want to know who committed these crimes and what causes these demographics to be the most targeted. 
