# JOUR405: Statistics for Journalists
## Midterm Exam - Spring 2025

Name: Nina Wilson 

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences. Start by loading the tidyverse and any other libraries you think you might need.

```{r}
library(tidyverse)
```


## Part 1: Restaurant Health Inspections (15 points)

You want to understand how restaurants in Montgomery County are performing on health inspections. The first dataset contains restaurant health inspection scores for restaurants in Montgomery County. The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible and the compliance score and grade. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv` and complete these tasks:
```{r}
health<-read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```


### Tasks:
1. Calculate the mean and standard deviation of compliance scores (5 points)

```{r}
health %>% summarize(mean=mean(compliance_score), sd=sd(compliance_score))

```

2. Create a histogram of the compliance scores with a vertical line showing the mean (5 points)
```{r}
health %>% 
  ggplot() +
  geom_histogram(aes(x=compliance_score), binwidth = 2)+
  geom_vline(aes(xintercept=mean(compliance_score)), color ="red", linetype="dashed", size=1,)
```

3. Write 2-3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What would be newsworthy about this distribution? What's the story here? (5 points).

The standard deviation is 5.826279 and the data is skewed left. Standard deviations that are closer to 1 are better because it shows how the data isn't as varied, so a standard deviation of 5.83 shows how the data is varied. A lot of schools got a compliance score of 100, which is really interesting and newsworthy I think. I would look into how these health inspections are run and whether they're just giving everyone 100's and good scores. 


## Part 2: High School Athletics (25 points)

You are reporting a story about high school sports participation in Maryland and want to see if there are differences between boys and girls. The second dataset shows participation numbers in high school sports across Maryland counties in 2024, broken down by sport and sex. Load the data from: `https://raw.githubusercontent.com/example/md_hs_sports_2024.csv` and complete these tasks:
```{r}
athletics<-read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_hs_participation.csv")
```


### Tasks:
1. Calculate the correlation between boys' and girls' participation (5 points)
```{r}
athletics %>%  summarize(correlation = cor(boys, girls, method="pearson"))
```

2. Add two columns called total and girls_pct using mutate(), with the total adding together boys and girls and girls_pct being the percentage of the total represented by girls participants. (5 points)
```{r}

athletics <- athletics%>% 
  mutate(
  girls_pct = ((girls/total)*100),
  total=(boys+girls)
)

```

3. Create a scatterplot showing this relationship, adding a line of best fit (5 points)

```{r}
athletics %>% 
  ggplot() +
  geom_point(aes(x=girls, y=boys)) +
  geom_smooth(aes(x=girls, y=boys), method="lm")
```
4. In 2-3 sentences, explain what the correlation coefficient and scatterplot reveal about equity in Maryland high school sports participation. How do you interpret the school districts that are below the line vs those that are above? Which school districts are most worth examining further, and why? (10 points)

The correlation is 0.982419, which is very close to 1, showing a strong relationship between girls and boys in sports. The line of best fit in the scatter plot confirms this because most of the points go off of the line and are near the line with the exception of 2 outliers. Schools that are above the line have a little bit more boys than girls in sports while schools below the line have less boys in sports. School districts that are most worth examining further are schools above the line because they have a more boys in sports than girls. 


## Part 3: Public Transit Ridership (20 points)

You are investigating public transit ridership in the Washington, D.C. area and want to understand the patterns of daily bus and rail ridership. The third dataset contains daily bus and rail ridership totals from WMATA for the past year. Load the data from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv and do the following:
```{r}
transit<-read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```


### Tasks:
1. Calculate the average bus and rail ridership and standard deviation using summarize() (5 points)
```{r}
transit %>% summarize(mean=mean(total), sd=sd(total))
```

2. Using the process you used in class, take a random sample daily ridership numbers and calculate the sample means and deviations for bus and rail. The number in the sample is up to you, but explain why you chose what you did. Compare this to the stats you generated in step 1. (5 points)
```{r}

```

3. Using group_by() and summarize(), calculate the means for bus and rail ridership for each weekday. Describe the overall pattern of ridership for bus and rail - which days stand out and why? Are there differences between bus and rail in the standard deviation values? (10 points)
```{r}

```



## Part 4: Maryland Car Theft Rates (20 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to analyze the data to find out which counties have the highest rates. The fourth dataset contains car theft statistics for Maryland counties in 2023 and population. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv` and complete the following tasks:
```{r}
thefts<-read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```


### Tasks:
1. Using mutate, add a column that calculates the rate of car thefts for each county - you need to choose the per capita rate (5 points)
```{r}
thefts <- thefts%>% 
  mutate(
    
```

2. Calculate the median car theft rate and the total number of car thefts statewide. Which counties have rates above the median, and what percentage of all car thefts occur in those counties? (5 points)
```{r}

```

3. Write 2-3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What's the lede of a story about your findings? (10 points)

## Part 5: Data Analysis Scenario (20 points)

You receive a tip that local emergency response times have gotten significantly worse over the past year. You obtain monthly data on response times for police, fire and ambulance calls.

Write 3-4 sentences (no code!) explaining:
1. What statistical measures would you calculate to verify this claim? (10 points)

I would add up the total response times per month and then do percent change for the months of last year and this year (and for the year as a whole) to see if there is a significant difference between response times over the past year. I would also try and see if I could find the reason why response times are getting worse, so I would look at the relationship between response times and temperature or weather or increasing crime rates, maybe more unemployment in the emergency response field... I would find the data for these other factors and find the correlation coeficiant to see if the relationship is strong or not. 

2. What visualizations would help readers understand the trends? (5 points)

I would make a histogram to show the frequency values of each month for each year, so two histograms total. I would also make a scatter plot with the line of best fit for whichever relationship seems the strongest. 

3. What additional context or data would you need to make this a complete story? (5 points)
For a complete story, I would look at which months are response times getting worse in and see if there's a relationship between temperature and response times or area and response times, is it increasing crime and response times. I would look at different factors that could be impacting response times the most. 

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your analytical choices
- Before submitting, clear your environment and run the entire notebook

Good luck!
