---
title: "HW11: Education Level and Survey Weighting"
author: "Nina Wilson
---

```{r setup, include=FALSE}
library(kableExtra)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# Introduction

In this assignment, you'll examine how survey weighting affects our understanding of voting intentions across different education levels, using the data we examined in class. You'll calculate both weighted and unweighted statistics, create visualizations, and reflect on the implications for reporting.

## The Data

The nonvoters dataset contains survey responses about voting intentions and behaviors from a national survey. The survey was conducted prior to an election and includes demographic information like education level.

```{r}
# Load the dataset
nonvoters_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/nonvoters_data.csv")

# Take a quick look at the data structure
glimpse(nonvoters_data)
```

### Key Variables

- `weight`: Survey weight assigned to each respondent
- `Q21`: Voting intention (1 = Yes, 2 = No, 3 = Unsure/Undecided)
- `educ`: Education level (College, Some college, High school or less)

## Task 1: Education Distribution

First, let's examine the distribution of education levels in our sample. Replace "REPLACE_ME" with the correct variable for education level.

```{r}

education_distribution <- nonvoters_data |>
  count(educ) |>
  mutate(percentage = n / sum(n) * 100) |>
  kable(digits = 1, col.names = c("Education Level", "Count", "Percentage (%)"))

education_distribution
```

## Task 2: Reflection Question

Why might education levels in survey samples often differ from the general population? What factors might cause certain education groups to be over or underrepresented?

Education levels in survey samples differ from the general population because some groups don't respond to surveys like these because they may not have the same access to technology. People with higher education would probably be overrepresented because college aged people understand the importance of data and have access to more technology most likely. 


## Task 3: Unweighted Analysis by Education

Now, let's calculate unweighted voting intentions by education level. This is what we would report if we didn't apply any weighting to our sample.

```{r}
# Calculate unweighted voting intentions by education
unweighted_by_education <- nonvoters_data |>
  # Filter out missing values
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  # Group by education and response
  group_by(educ, Q21) |>
  # Count responses
  summarize(count = n(), .groups = "drop_last") |>
  # Calculate percentages
  mutate(total = sum(count),
         percentage = count / total * 100) |>
  ungroup()

# Create a more readable format with voting intentions as columns
unweighted_educ_summary <- unweighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

kable(unweighted_educ_summary, digits = 1, caption = "Unweighted Voting Intentions by Education Level")
```

## Task 4: Reflection Question

Based just on this unweighted analysis, what headline might you write for a news story about education and voting intentions?

College students and graduates lead the voting population

## Task 5: Weighted Analysis by Education

Next, let's apply survey weights to see how this changes our results. Instead of just counting responses, we'll sum the weights for each group. Replace "REPLACE_ME" with the appropriate weight variable

```{r weighted-by-education}

weighted_by_education <- nonvoters_data |>
  # Filter out missing values
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  # Group by education and response
  group_by(educ, Q21) |>
  # Sum the weights instead of counting
  summarize(weighted_count = sum(weight), .groups = "drop_last") |>
  # Calculate weighted percentages
  mutate(weighted_total = sum(weighted_count),
         weighted_percentage = weighted_count / weighted_total * 100) |>
  ungroup()

# Create a more readable format
weighted_educ_summary <- weighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = weighted_percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

kable(weighted_educ_summary, digits = 1, caption = "Weighted Voting Intentions by Education Level")
```

## Task 6: Reflection Questions

1. How did the percentages change after applying weights? Which education group showed the biggest changes?

After applying the weights, the percentages only changed slightly with a drop or add of a couple percentage points. The group with the biggest change was high school or less with a drop in those who voted no and added in those that voted unsure. 

2. Why might the weighted results be considered more accurate than the unweighted results?
The weighted results might be considered more accurate than the unweighted results because its meant to reflect the general population by correcting for overrepresentation and underrepresentation of certain groups. 


## Task 7: Comparison of Weighted vs. Unweighted Results

Let's create a direct comparison table to see the differences more clearly.

```{r}

comparison <- unweighted_educ_summary |>
  inner_join(weighted_educ_summary, by = "educ", suffix = c("_unweighted", "_weighted")) |>
  mutate(
    # Calculate the differences between weighted and unweighted percentages
    yes_diff = `Yes (%)_weighted` - `Yes (%)_unweighted`,
    no_diff = `No (%)_weighted` - `No (%)_unweighted`,
    unsure_diff = `Unsure (%)_weighted` - `Unsure (%)_unweighted`
  ) |>
  # Select just the columns we want to display
  select(educ, yes_diff, no_diff, unsure_diff) |>
  rename(
    "Education Level" = educ,
    "Yes (% point diff)" = yes_diff,
    "No (% point diff)" = no_diff,
    "Unsure (% point diff)" = unsure_diff
  )

kable(comparison, digits = 1, caption = "Difference Between Weighted and Unweighted Results (percentage points)")
```

## Task 8: Reflection Question

Which education group shows the largest differences between weighted and unweighted results?

People who had an education of high school or less had the biggest change between weighted and unweighted results with a change of -2.7 percentage points. High school or less had the biggest percentage change in general with all of the catergories being over 1. 

## Task 9: Visualization

Visualizations can help us see the differences more clearly. Let's create a bar chart comparing weighted and unweighted "Yes" responses by education level. Replace "REPLACE_ME" with the correct variable name

```{r}
educ_viz_data <- bind_rows(
  # Unweighted data
  unweighted_by_education |> 
    filter(Q21 == 1) |>  # Only "Yes" responses (Q21=1)
    mutate(Type = "Unweighted") |>
    select(Type, educ, percentage),
  
  # Weighted data - 
  weighted_by_education |> 
    filter(Q21 == 1) |>  # Only "Yes" responses
    mutate(
      Type = "Weighted",
      percentage = weighted_percentage
    ) |>
    select(Type, educ, percentage)
)

# Create a grouped bar chart
ggplot(educ_viz_data, 
       aes(x = educ, y = percentage, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(
    title = "Weighted vs. Unweighted 'Yes' Responses by Education",
    subtitle = "Q21: Do you plan to vote in the November election?",
    y = "Percentage (%)",
    x = "Education Level"
  ) +
  scale_fill_manual(values = c("Unweighted" = "#619CFF", "Weighted" = "#F8766D")) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## Task 10: Reflection Questions

Does the visualization make it easier to see the differences between weighted and unweighted results? Why or why not?

The data visualization makes it a little easier to see the differences between weighted and unweighted results because of the percentages on top of the bars. I think it would work better if the differences were bigger or if the units were smaller so that visually we can see the difference. 

## Task 11: Summary

Based on your analysis of weighted and unweighted results by education level, write a brief (2-3 paragraph) journalistic summary of what you found. You should try to address:

1. How education level relates to voting intentions
2. How weighting affected your understanding of this relationship
3. What this means for interpreting polling results in news reporting

The higher the education level, the more likely the person votes. Around 93% of people who finished college planned on voting while around 85% of people who finished some college are likely to vote. Less than 75% of people with a high school education or less plan on voting. 

Weighting showed the more accurate representation of the survey, kind of showing how much more people have just a high school education versus a college education. A high school education or less is more common than a college education because it's free, and mandatory for people to do. College is more of a commitment and not as accessible to different demographics with money and time. 

For interpreting polling results in news reporting, it's important to look at a specific community as a whole and its poll results but it's just as important to obtain the weighted results to compare the community's results to society as a whole. 

## Task 12: Final Reflection Questions

1. Why is it important for journalists to understand survey weighting when reporting on polls?
It's important for journalists to understand weighting when reporting on polls because without weighting, data could misrepresent the public and the data could be skewed.

2. How might the differences between weighted and unweighted results affect how you would report on this data?
The differences between weighted and unweighted results impact the different interpretations a reporter could take. One reporter could take a interpretation of a high voter turnout for people with a high school education when in the general public, this is not the case. 

3. What additional information would you want to know about how the weights were calculated before using this data in a news story?
I would want to know more about race and gender within these polling results. Like what percentage of college educated voters are white men and white women, how many are black men and black women? I think this would lead to deeper questions and make for a much interesting story on who we see voting and why that may be. 
